---
title: "1986-2018 Ecuador Anthropometric Database"
author: "Philippe Belmont Guerrón"
date: "2023-10-30"
output: html_document
always_allow_html: true
bibliography: references.bib
---

```{r setup, include=FALSE}
library(kableExtra)
library(tidyverse)
knitr::opts_chunk$set(echo = F)
db <- readRDS("antro1986_2018_n5.rds")
options(scipen = 8)
```


This dataset integrates more than three decades of public health databases
focusing on the Ecuadorian population. It combines individual data from six
distinct datasets, encompassing a total of `r nrow(db)` individuals and `r ncol(db)` variables.
This consolidated database includes anthropometric measurements, along with
socio-economic characteristics of both households and individuals. Within this
database, there are `r sum(as.numeric(db$tpind==2))` instances of child-mother
pairs, involving children under the age of 5, as well as
`r sum(as.numeric(db$tpind==1))` unique mothers. Additionally, the dataset
comprises `r sum(as.numeric(db$tpind==3))` other individuals, primarily
consisting of individuals over the age of 5 and Women in Reproductive Age (WRA).

# Sources

The dataset was created using the data collected from the following national
health surveys:

-   DANS 1986: Diagnóstico de la situación alimentaria, nutricional y de salud
    de la población ecuatoriana menor de cinco años [@wilmab.freire1988],
-   ENDEMAIN 1999: Encuesta Demografía y de Salud Materna e Infantil
    [@catón2005],
-   ENDEMAIN 2004: Encuesta Demografía y de Salud Materna e Infantil
    [@ordóñez2005],
-   ENSANUT 2012: Encuesta Nacional de Salud y Nutrición [@freire2014],
-   ECV 2014 : Encuesta Nacional de Condiciones de Vida [@inec2014],
-   ENSANUT 2018 : Encuesta Nacional de Salud y Nutrición 2018 [@inec] .

# Available data

The total sampled population of children under 5 years old remained below 10,000 from
1986 to 2012. Subsequently, there was a notable increase in the 2014 and 2018
surveys. Additionally, anthropometric data for other population groups has shown
a steady increase from 2004 to 2018.

```{r cars, echo=F}
tb1 <- db %>% group_by(idenc,tpind) %>% summarise(n=n(),.groups = 'drop') %>%
  spread(.,idenc,n)
 tb1$Total <- rowSums(tb1[,-1])
 tb1 <- rbind(tb1,colSums(tb1))
 tb1$tpind <- c("Mothers","Child<5yo","Other","Total")
 names(tb1)[1] <- "Group"
 tb1 <- as.data.frame(tb1)
kbl(tb1, align="l",escape=F,
    caption = "Total of observations",
    booktabs = TRUE) %>%
  row_spec(3,hline_after = T) %>%
  kable_styling(full_width = T)

```

# Anthropometric data

Anthropometric indicators were calculated for three age groups separately: 0 to
under 5 years, 5 to 19 years and 20 to 59 years. Standardized table and
calculation methodology employed in the first two groups were conducted with
igrowup (Stunting:Haz\<-2, OW:2\>zBMI\>=3, OB:zBMI\>3) and who2007
(OW:1\>zBFA\>=2, OB: zBFA \>2) package provided by the WHO [5]. For the third
group from 20 to 59 years, BMI was calculated and cut-offs published by the WHO
were applied to evaluate overweight and obesity (OW:25\>=BMI\>30, OB:BMI\>=30).
For each survey the availability of pregnancy state (*embrz*) provided a filter
to exclude all women in pregnancy from anthropometrics status estimation.
Anthropometric indicators were computed separately for three distinct age
categories: 0 to under 5 years, 5 to 19 years, and 20 to 59 years. The
standardized table and calculation approach employed for the first two age
groups were executed using the "igrowup" package (Stunting:Haz\<-2,
OW:2\>zBMI\>=3, OB:zBMI\>3) and the "who2007" package (OW:1\>zBFA\>=2, OB: zBFA
\>2), both provided by the World Health Organization [@deonis2006].

For the third age group (20 to 59 years), Body Mass Index (BMI) was computed,
and the BMI thresholds published by the World Health Organization were used to
determine overweight and obesity status (Overweight: 25 \<= BMI \< 30, Obesity:
BMI \>= 30).

In each survey, the availability of pregnancy status (*embrz*) was utilized as a
filter to exclude all pregnant women from the estimation of anthropometric
status.

# Complex Survey design

For each individual database, the original survey design was taken into account,
incorporating sampling weights (*pw*), strata (*strata*), and primary sampling
units (*psu*) into the final consolidated database. The sampling design ensured
that representativeness was achieved at the provincial level for the most recent
surveys (ENDEMAIN 2004, ENSANUT 2012, ENSANUT 2018), and at the geographical domains (Highlands, Coast, Amazon, Quito and Guayaquil).

Over the years, the size of the surveys grew substantially, covering a broader
geographical area. Notably, in 1986, surveys did not encompass the Amazon region
or the Galapagos Islands due to technical constraints.

In the 2018 survey, there was no self-representation available for the two main
cities in the country: Quito and Guayaquil.

```{r}

tb2 <- db %>% group_by(subreg,idenc) %>% summarise(n=n(),.groups = 'drop') %>%
  spread(.,idenc,n)
tb2 <- rbind(tb2,colSums(tb2,na.rm = T))
tb2[is.na(tb2)] <- 0
tb2[11,]<- tb2[1,]+tb2[8,]
tb2[12,]<- tb2[3,]+tb2[9,]
tb2 <- tb2[c(8,1,11,2,12,3,9,4:7,10),]


tb2$subreg <- c("Quito","Other","Subtotal","Subtotal",
                "Guayaquil","Other","Subtotal","Subtotal",
                "Subtotal", "Subtotal", "Galapagos" ,"Total")
tb2 <- cbind(Region=c("Sierra"," "," "," ","Costa"," "," "," ",
                "Amazonia"," ","Galapagos","Total"),
             Area=c("Urban"," "," ","Rural","Urban"," "," ","Rural",
                "Urban","Rural","Galapagos","Total"),
             tb2)
names(tb2)[names(tb2)=="subreg"] <- "Area"

tb2 <- as.data.frame(tb2)
kbl(tb2, align="l",escape=F,
    caption = "Observations per Dominion",
    booktabs = TRUE) %>%
  row_spec(c(4,8,10,11),hline_after = T) %>%
  row_spec(c(3,7,9),extra_latex_after = "\\cline{2-9}") %>%
  row_spec(c(2,6),extra_latex_after = "\\cline{3-9}") %>%
  kable_styling(full_width = T)


```

# Imputation of data

In the 1986 dataset, missing day of birth values for Women in Reproductive Age
(WRA) and children were imputed using the methodology outlined in the DHS
approach (refer to: [@rutstein2006]), with months being used for missing values.
In subsequent surveys, no additional imputation was conducted due to the
comprehensive nature of the data and the limited impact of missing values.

For the 2012 survey, anthropometric data was evaluated independently of the
reproductive survey. Consequently, a portion of cases did not contain
information about the number of children (2200 cases out of 21296 cases),
although this constituted less than the 10% threshold that would warrant
applying an imputation method. An imputation technique involving chained
equations was employed, considering variables such as age groups, region,
education, and crowding. This regression-based algorithm iteratively filled in
missing values across multiple variables. The approach followed a chained
equations strategy, utilizing fully conditional specification (FCS) in the
prediction equations.

# Household material well being index

This index was created by utilizing unmet essential requirements, acting as a surrogate measure for poverty, with a particular focus on the tangible dimension. To account for the broad time frame of the data, the elements were consolidated to encompass data from diverse surveys. The Unmet Basic Needs (UBN) index offers a means to establish a metric for well-being grounded in critical physical necessities related to housing and infrastructure, especially in situations where data is limited (see: [@hammill2009]). The components integrated into this index (referred to as *ics* in the database) were:


-   Crowding more than three persons per room;
-   Housing: living in a house made of irregular materials considering floor,
    walls and house type;
-   Sanitation : not having an indoor flush toilet;
-   Cooking fuel: not using gas or electricity as cooking energy;
-   Water: not having an indoor tubed water;
-   Waste disposal: not accessing to motorized garbage disposal.

Sum of each component (unweigthed) and classification per 2 points category was
calculated to build the final index.

# Reference
